Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-04 tagless/tagless-final.tex
Done with Section 4.3; started with Section 4.4 (simplified PE code shown)

\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}

@@

It is straightforward to create an algebraic data type, say in OCaml, to
represent object terms such as those in Figure~\ref{fig:object}.
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
\begin{code}
type var = VZ | VS of var
type exp = V of var | B of bool
         | L of exp | A of exp * exp | ...
\end{code}
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}

@@

let rec eval0 env = function ...
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

@@

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions.
\begin{code}
type u = UB of bool | UA of (u -> u) | ...
\end{code}
We can then write a typed interpreter
\begin{code}
let rec eval env = function ...
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f ->
               f (eval env e2)
\end{code}
whose inferred type is |u list -> exp -> u|. Now we can evaluate
our sample |test1| term.
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions.
\jacques{but now this smaller language does not match anymore what
we wrote earlier.  I think it is better to use a small language here,
but i don't quite know how to say so}
\begin{code}
let varZ (x,env)    = x
let varS vp (x,env) = vp env
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code}
These functions comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment.
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
The result has no tags---the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as |testf3| below gives a type error rather than
a run-time error.
\begin{code}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.

@@

In the sequel, we encode binding in object programs using higher-order
abstract syntax \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices.

@@

We have also implemented this compiler in Haskell. Since Haskell
has no (convenient, typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
Whereas our representation of object terms uses higher-order
abstract syntax, our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \textsf{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code given above. The
accompanying code gives the full details.

@@

To enable these simplifications, we delay binding-time analysis (BTA)
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1| data structure: applying the type constructor
|P1| to a function type |a -> b| should yield one of
\begin{code}
data P1 (a -> b) = S1 (P1 a -> P1 b)
                 | E1 (C (a->b))

@@

\subsection{GADT -- not quite tagless PE}
\label{S:PE-GADT}

Two common ways of providing typecase in Haskell is by using either
GADT or type-class functions
(implemented via typeclasses with functional dependencies)
\citep{oliveira-typecase}.  These
methods are equivalent, and here we show the GADT; please see |incope1.hs|
in the accompanying source code for the latter.

We introduce the following GADT
\begin{code}
data P t where
  VI :: Int  -> P Int
  VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a->b)
  E  :: C t -> P t
\end{code}
with four data constructors. The constructor |VF| stands out as its
operand has type |P a -> P b| rather than |a -> b| as one may
have expected if |P t| were parametric in |t|. As before, we define a
function to extract the future-stage computation from the |P t| value:
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}

We use the method |lam| of the |C| intrepreter to compile the argument
of |lam| into a code value. We may now make |P| an instance of
Symantics and implement the partial evaluator (the
implementations of |mul|, |leq|, |if_| and |fix| are elided) as follows.
\begin{code}
instance Symantics P where
int x  = VI x
bool b = VB b

add e1 e2 = case (e1,e2) of
             (VI n1,VI n2) -> VI (n1 + n2)
             _ -> E (add (abstr e1) (abstr e2))
lam = VF
app ef ea = case ef of
             VF f -> f ea
             E  f -> E (app f (abstr ea))
\end{code}

The implementations of |int|, |bool|, and |add| are similar to the
above. The interpretation of |lam f| merely incorporates the
HOAS function |f| into the result. At any point we can convert it to a
code value, if we must. By delaying such a conversion, we get the
opportunity to apply |f| to a concrete argument. When evaluating
|app ef ea|, we check to see of the first argument is such a delayed
lambda-expression. If it is, we apply the incorporated |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example |testpowfix| in the next subsection). If |ef| is a
dynamic value, we residualize.

Although GADTs solve our problem, the solution is not quite
satisfactory. First of all, it cannot be ported to MetaOCaml as GADTs
are not supported there. Second, the tagging problem reappeared, in
the implementation of |app| above. The datatype |P t| has four data
constructors; pattern-matching in |app| however uses only two of
them, |VF| and |E|. It is obvious, one may say, that the constructors
|VI| and |VB| just cannot occur, because the corresponding values
cannot be function types. What is obvious to us is not obvious to
the compiler. The case matching above looks \emph{partial}.  While this
may seem like a minor point, 
this is precisely the main point of
tag elimination. In a typed tagged interpreter there are many
pattern-matching forms that looks partial but are total in
reality. The goal is to make this totality \emph{syntactically}
apparent.

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in Section~\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = S (repr_pe t) (C t) | E (C t)
\end{code}
where |repr_pe| is a type function, defined inductively
% inductively because P below depends on repr_pe
as 
\begin{code}
repr_pe Int    = Int
repr_pe Bool   = Bool
repr_pe (a->b) = P a -> P b
\end{code}
Although we can use typeclasses to define this type function
in Haskell, that is not portable to MetaOCaml. We notice however that
the type case alternatives above are all associated with existing
methods of |Symantics|, and so the case analysis is already present. 
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml for
portability, we recall from Section~\ref{encoding} that the |repr| type
had two arguments |type ('c,'dv) repr|. We add an extra argument,
|'sv|, which is the result of applying the |repr_pe| function above to the
type |'dv|. Figure~\ref{fig:ocaml} shows the new signature |Symantics|.
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig

@@

  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
\label{fig:ocaml}
\end{figure*}

The interpreters |R|, |L| and |C| above only relied on the last
argument of the |repr| type, which is unchanged. And so is their
implementation; the newly added argument |'sv| will be phantom.
For example, we would now write the interpreter |C| as
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
When writing the partial evaluator however, we make use of the
type argument |'sv|. Here is the corresponding instance of
|Symantics|:

\begin{code}
module P = struct
  type ('c,'sv,'dv) repr = 
        {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x
  let pdyn x = {st = None; dy = x}

@@

  let build cast f1 f2 = fun e1 -> fun e2 -> match (e1,e2) with
                   ({st = Some n1}, {st = Some n2}) -> cast (f1 n1 n2)
                 | _ -> pdyn (f2 (abstr e1) (abstr e2))
  let buildsimp_monoid cast e f1 f2 = fun e1 e2 -> match (e1,e2) with
                 | ({st = Some e'}, _) when e = e' -> e2
                 | (_, {st = Some e'}) when e = e' -> e1
                 | _ -> build cast f1 f2 e1 e2
  let buildsimp_ring cast zero one f1 f2 = fun e1 e2 -> match (e1,e2) with
                 | ({st = Some e'}, _) when e' = zero -> e1
                 | (_, {st = Some e'}) when e' = zero -> e2
                 | _ -> buildsimp_monoid cast one f1 f2 e1 e2

  let add e1 e2 = buildsimp_monoid int 0 R.add C.add e1 e2
  let mul e1 e2 = buildsimp_ring int 0 1 R.mul C.mul e1 e2
  let leq e1 e2 = build bool R.leq C.leq e1 e2
  let if_ eb et ee = match eb with
           {st = Some b} -> if b then et () else ee ()
         | _ -> pdyn (C.if_ (abstr eb) 
                          (fun () -> abstr (et ()))
                          (fun () -> abstr (ee ())))
  let lam f = {st = Some f; 
               dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with
        {st = Some f} -> f ea
      | _ -> pdyn (C.app (abstr ef) (abstr ea))

  let fix f = 
    let fdyn = C.fix (fun x -> abstr (f (pdyn x))) in
    let fdynn e = pdyn (C.app fdyn e.dy) in
    {st=Some (function {st = Some _} as e -> 
               let rec self n = 
                (match n.st with Some _->app (f (lam self)) n 
                                | _ -> fdynn n) in 
               app (f (lam self)) e
             | e  -> fdynn e);
    dy = fdyn }
end;;
\end{code}

The type |repr| here literally expresses the type equation at the
beginning of this section. The function |abstr|, as in the previous
subsection, extracts the future-stage computation from the result of
partial evaluation.  The function |pdyn| does the converse. As in
section~\ref{S:PE-problem}, we build future-stage computations at the
same time as we
build present-stage ones, if any. Therefore, we avoid the need for a
polymorphic `lift' function. The implementations of
|add| and |mul| are a bit more sophisticated, taking advantage 
respectively of the implicit monoid and ring structure of the 
underlying domain.
The function |build| checks to see if
its two arguments |e1| and |e2| are static expressions. If so, the
corresponding static computation (with the help of the intrepreter
|R|) is performed. The function |buildsim_monoid| also checks if one argument
is static and is the neutral element of the monoid structure.
|buildsimp_ring| also worries about the full ring structure rather than
just the monoid structure.  These are put here to illustrate how we 
would go about writing general algebraic simplification routines that
would mimic various compiler optimizations, and these would be quite
beneficial in a large language with more base types.

>>>>

\hyphenation{meta-language meta-languages Meta-OCaml meta-cir-cu-lar type-case}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage[override]{cmtt}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}

@@

It is straightforward to create an algebraic data type, say in OCaml, to
represent object terms such as those in Figure~\ref{fig:object}.
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
\begin{code}
type var = VZ | VS of var
type exp = V of var | B of bool
         | L of exp | A of exp * exp
\end{code}
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}

@@

let rec eval0 env = function
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code}

@@

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions.
\begin{code}
type u = UB of bool | UA of (u -> u)
\end{code}
We can then write a typed interpreter
\begin{code}
let rec eval env = function
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f ->
               f (eval env e2)
\end{code}
whose inferred type is |u list -> exp -> u|. Now we can evaluate
our sample |test1| term.
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions.
\jacques{but now this smaller language does not match anymore what
we wrote earlier.  I think it is better to use a small language here,
but i don't quite know how to say so}
\begin{code}
let varZ env        = fst env
let varS vp env     = vp (snd env)
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code}
These functions comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment.
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
The result has no tags---the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as |testf3| below gives a type error rather than
a run-time error.
\begin{code}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.

@@

In the sequel, we encode binding in object programs using higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices.

@@

We have also implemented this compiler in Haskell. Since Haskell
has no (convenient, typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \textsf{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code given above. The
accompanying code gives the full details.

@@

To enable these simplifications, we delay binding-time analysis (BTA)
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1| data structure: applying the type constructor
|P1| to a function type |a -> b| should yield one of
\begin{code}
data P1 (a -> b) = S1 (P1 a -> P1 b)
                 | E1 (C (a -> b))

@@

\subsection{Eliminating tags from typecase}
\label{S:PE-GADT}

Two common ways to provide typecase in Haskell are
GADTs and type-class functional dependencies
\citep{oliveira-typecase}.  These
methods are equivalent, and here we show the GADT way; |incope1.hs|
in the accompanying source code shows the latter.

We introduce a GADT with four data constructors.
\begin{code}
data P t where
  VI :: Int  -> P Int
  VB :: Bool -> P Bool
  VF :: (P a -> P b) -> P (a -> b)
  E  :: C t -> P t
\end{code}
The constructors |VI|, |VB|, and |VF| build static terms (like |S0|
in~\S\ref{S:PE-lift}), and |E| builds dynamic terms (like |E0|).  However,
the type |P t| is no longer parametric in~|t|: the constructor |VF| takes an
operand of type |P a -> P b| rather than |a -> b|. As before, we define a
function to extract a future-stage computation from a value of type |P t|.
\begin{code}
abstr :: P t -> C t
abstr (VI i) = int i
abstr (VB b) = bool b
abstr (VF f) = lam (abstr . f . E)
abstr (E x)  = x
\end{code}
The cases of this function |abstr| are type-indexed.  In particular, the |VF f|
case uses the method |lam| of the |C| interpreter to compile~|f|.

We may now make |P| an instance of
|Symantics| and implement the partial evaluator as follows. We elide
|mul|, |leq|, |if_|, and |fix|.
\begin{code}
instance Symantics P where
  int x  = VI x
  bool b = VB b
  add (VI n1) (VI n2) = VI (n1 + n2)
  add e1 e2 = E (add (abstr e1) (abstr e2))
  lam = VF
  app (VF f) ea = f ea
  app (E f)  ea = E (app f (abstr ea))
\end{code}
The implementations of |int|, |bool|, and |add| are like
in~\S\ref{S:PE-problem}.
The interpretation of |lam f| just wraps the
HOAS function |f| into the result. We can always compile |f| to a code value,
but we delay it to apply |f| to concrete arguments. The interpretation of
|app ef ea| checks to see if |ef| is such a delayed
HOAS function |VF f|. If it is, we apply |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
dynamic value |E f|, we residualize.

This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml as GADTs are unavailable there.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in |app|
above: the type |P t| has four constructors, of which pattern\hyp
matching in |app| uses only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
Indeed the compiler could reason thus, but it may not; GHC for one
issues warnings.  Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but are total in reality. The goal is to make this totality
\emph{syntactically} apparent.

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code}
repr_pe Int      = Int
repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new |Symantics| signature.
\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig

@@

  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
  val if_ : ('c,bool,bool) repr -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)

end
\end{code}
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports partial evaluation}
\label{fig:ocaml}
\end{figure*}

\begin{figure}
\begin{floatrule}
\begin{code}
module P = struct

  type ('c,'sv,'dv) repr = {st: 'sv option;
                            dy: ('c,'dv) code}

  let abstr {dy = x} = x
  let pdyn x = {st = None; dy = x}

@@

  let add e1 e2 = match e1, e2 with
  | {st = Some 0}, e | e, {st = Some 0} -> e
  | {st = Some m}, {st = Some n} -> int (R.add m n)
  | _ -> pdyn (C.add (abstr e1) (abstr e2))

  let if_ eb et ee = match eb with
  | {st = Some b} -> if b then et () else ee ()
  | _ -> pdyn (C.if_ (abstr eb) 
                     (fun () -> abstr (et ()))
                     (fun () -> abstr (ee ())))

  let lam f =
  {st = Some f; 
   dy = C.lam (fun x -> abstr (f (pdyn x)))}

  let app ef ea = match ef with
  | {st = Some f} -> f ea
  | _ -> pdyn (C.app (abstr ef) (abstr ea))

  let fix f = 
    let fdyn = C.fix (fun x -> abstr (f (pdyn x)))
    in let rec self = function
       | {st = Some _} as e -> app (f (lam self)) e
       | e -> pdyn (C.app fdyn (abstr e))
       in {st = Some self; dy = fdyn}

end
\end{code}
\end{floatrule}
\caption{Our partial evaluator (eliding \texttt{mul} and \texttt{leq})}
\label{fig:pe}
\end{figure}

The interpreters |R|, |L| and~|C| above only use the old
type arguments |'c| and~|'dv|, which are treated by the new signature
in the same way.  Hence all that needs to change in these interpreters
to match the new signature is to add a phantom type
argument~|'sv| to~|repr|.
For example, the compiler |C| now begins
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
  (* the rest is the same *)
\end{code}
In contrast, the partial evaluator~|P| relies on the type argument |'sv|.

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| literally expresses the type equation for |repr_pe| above.
The function |abstr|, as in~\S\ref{S:PE-GADT},
extracts a future-stage code value from a result of
partial evaluation.  Conversely, the function |pdyn| coerces a
code value into a partial\hyp evaluation result. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
static ones, if any, to avoid
polymorphic lift. The implementations of
|add| and |mul| are a bit more sophisticated, taking advantage 
respectively of the implicit monoid and ring structure of the 
underlying domain.
The function |build| checks to see if
its two arguments |e1| and |e2| are static expressions. If so, the
corresponding static computation (with the help of the intrepreter
|R|) is performed. The function |buildsim_monoid| also checks if one argument
is static and is the neutral element of the monoid structure.
|ring| also worries about the full ring structure rather than
just the monoid structure.  These are put here to illustrate how we 
would go about writing general algebraic simplification routines that
would mimic various compiler optimizations, and these would be quite
beneficial in a large language with more base types.
